Bayesian exponential random graph modeling of KWL Burial network

### 2020-12-17 analysis/paper/paper.Rmd
edit intro, result, and conclusion section

<!-- two types of network analysis -->
Network analysis has been increasingly applied by archaeologists in recent years to deal with past interactions and explore the underlying mechanisms. There are two common approaches to characterize network properties at two distinct scales: node/edge level and graph level [@Peeples2019]. Node level focuses on the role of node in a network, such as centrality, representing the individual influence or social prominence in a group, while graph level assesses the whole network attributes, such as density, clustering in a network, to generalize relationship patterns [@Mills2017; @Peeples2019]. By quantifying those network properties, archaeologists can answer a wide range of research questions. Examples includes exploring the political centralization in the Kofun period in Japan through the hierarchical communication network constructed by prestige goods [@Mizoguchi2013], the investigation of long term inter-site relationships from the Epipalaeolithic to the early Neolithic in the Near East according to trade items [@Coward2013]. Regarding burials, @Sosna2013 examined spatial pattern of burials from the Early Bronze Age in Rebesovice with two hypothesized networks constructed according to cultural and chronological similarity between burials. Recently, complex network modeling evaluates networks at both node and graph level through simulations of particular processes and statistically testing the formation of network properties [@Brughmans2013; @Brughmans2018; @Freeman2004; @Salvini2010]. Such application includes simulations and testing food exchange modes for Ancestral Pueblos on the aggregation of households in the American Southwest simulation [@Crabtree2015], or exploring the diffusion of fired bricks across Hellenistic Europe by comparing similarity networks of sites with random networks. Other examples show the assessment of hunter-gatherer exchange networks structure across the Kuril Islands using bootstrap simulation based on ceramic composition [@Erik2015]. This paper will use a novel Bayesian modeling approach to investigate the formation of relationship between burials in northeastern Taiwan to explore the indirect impact of foreign contacts on social organization.

@@

<!-- archaeological specific case study -->
In this study, we use a pericolonial model to explore the changes in social inequality at Kiwulan, an Iron Age site in northeastern Taiwan, which covers the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century [@Chen2007]. We assume that the social changes at Kiwulan would be supported by chronological differences in the structure of burial networks. We examine burial networks before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) does the observed burial data after the start of European presence resemble the network of our hypothesized model? (2) did European colonial activities in 17th century Taiwan result in the emergence of social inequality in an Indigenous society in ways that can be detected by burials networks? (3) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of burials in understanding the indirect effects of a colonial presence on Indigenous groups. This is important to understand the reactions of Kiwulan residents in the context of indirect impacts of European colonization.

@@

<!-- current approaches to network analysis: important to show how what we are doing is amazing and new and not done before by other archaeologists. We need a sentence or two to say what do archaeologists usually do for network statistics... and make it clear that we are doing something new -->
Exponential Random Graph Models (ERGMs) are an important family of statistical models for networks that allows direct modeling for the formation of edges, or ties, between nodes [@Robins2007]. The assumption is that possible ties in a network are random variables and dependent on actor variables or the presence or absence of other ties [@Robins2007]. ERGMs differs from traditional network approaches in that edges are only formed according to their corresponding predictors separately without taking into account the interactions among edges [@Morris2008]. In ERGMs, networks are viewed as dependent variables, the formation of a tie is also influenced by network dependencies, and the attributes of nodes/edges [@Snijders2011]. For example, nodes with similar attributes are more likely to form a relationship, such as friendship between people with the same hobby. Ties form a small structure in a network called a graph configuration, that describes the form of dependence, such as reciprocity (relationship between two actors), transitivity or clustering (relationship between two actors through a shared third actor), homophily (relationship between actors with a similar attribute), or popularity (actors have many relationships with others) [@Robins2007; @Snijders2006; @Morris2008]. Those configurations represent the structure or the property of a network and can be expressed by network statistics. We can model those network statistics as direct functions of ties, where the probability of occurrence of ties can be specified and hypothesized [@Morris2008; @Robins2007]. By specifying the forms of configurations, we can build a hypothesis-based model to generate a distribution of random networks that represent our model. Such distribution consists of a large number of possible networks that enables statistical inference and comparison with an observed network [@Robins2007]. This helps us understand whether an observed network shows significantly more or less of a property of interest than the random networks generated from our model assumptions.

<!-- ERGM in archaeology, and limitations --> 
To date, ERGM has only been used once in archaeology, by @Amati2019 who reconstructed three different networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore their interaction mechanisms. These network models evaluated different hypotheses about the importance of three variables, proximity, inter-cultural items, and pottery types, on the presence of hub sites they observed. They found that the presence of hubs can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively. They also pointed out some limitations of ERGMs, such as static outcomes and sensitivity to missing data. In addition, it is difficult in ERGMs to estimate model parameters and interpret the result due to intractable likelihood normalizing constants and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a function of the model parameter for making probability distributions integrate to one, which becomes harder to compute with larger set of networks [@Caimo2017bayesian]. This is also termed "doubly intractable" since both the likelihood normalizing constant and the marginal likelihood (the evidence of the posterior) are hard to derive [@Caimo2013; @Lyne2015]. Model degeneracy is another issue that probability models tend to overestimate a small number of extreme graphs by assigning too much weight, such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. One solution to these limitations is available by implementing ERGMs in a Bayesian framework.

@@

<!-- comparison between two networks using bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To understand the robustness of comparison between two networks, we used vertex bootstrap to cross-validate the results based on Bayesian network analysis. Vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) of a network to quantify standard errors in the network statistics of interest [@Chen2019; @Snijders1999]. This enables the evaluation of uncertainty for networks and tests the difference between networks by examining their confidence intervals for the network population, where networks with unoverlapped intervals represent a significant difference. We set the resampling size to 1,000 networks to compute endogenous network statistics, including density, transitivity, and popularity for our two networks. The result shows a significant difference in network popularity between the two networks, which is consistent with our finding of negative popularity in the pre-European period and positive popularity in the post-European period using the Bayesian approach. For density and transitivity, both 95% confidence intervals show some level of overlapping that indicates no significant difference. This pattern is also similar to our results of Bayesian modeling, where both networks show similar positive or negative effects with some level of differences.

Similarly, vertex bootstrap allows statistical testing for multiple networks, such as a t-test method. We conducted a t-test on the distributions of network statistics between two networks. The results show that there are significant differences in all network statistics. This supports the presence of anthropologically meaningful differences in the pre-and post-European networks. Moreover, we tested the sample size effect by removing nodes at certain percents (5-30%) and compare their confidence intervals for both networks using the same vertex bootstrap procedure. The results show a consistent difference between the two networks that shows they are robust to the different sizes. In general, the cross-validation of results from vertex bootstrap and Bayesian modeling illustrates the robust comparison between two networks without sample size effects.

@@

<!-- summary of the research, contribution and methodological implication-->
In this paper, we presented a naval approach for studying burial relations using ERGM network analysis within a Bayesian framework to rigorously examined social changes in a European percicolonial context in northeastern Taiwan. We tested the hypotheses of chronological changes in interburial networks at Kiwulan with the evaluation of both endogenous and exogenous network effects. Our results provide a meaningful explanation for the relationship between burials that share similar prestige goods. Before European contact in the 17th century, the network properties has a tendency of more closed burial relationships with ritual elements as the key formation mechanism. After the European arrival, the network has a tendency of centralization relative to the rarity of goods. This aliens with changes in social complexity from a more corporate to a more network mode society that stress different strategies for achieving power and different degrees of social inequality. Based on the burial data, along with historical records and ethnography, we are able to understand the processes of changes in Indigenous societies in the periphery of European colonial bases. This study demonstrates not only the pericolonial effect on local Indigenous societies that have not well understood yet, but also the agency of local people to manipulate colonial power instead of passive recipients of trade goods.

This case study highlights the methodological improvement that Bayesian inference on ERGM can inform and enhance studies in rational data in archaeology. A Bayesian framework can reduce the effects of small sample size or missing data commonly present in archaeological data by incorporating prior information and MCMC estimations. Moreover, probabilistic inference provides robust results of interrelations that are not statistically tested by traditional network approaches. Bayesian network modeling can be applied to a wide range of archaeological data to examine the formation of relationships. This allows in-depth discussion of dynamic processes of relationship formation and their underlying factors that are informative of historical trajectories or socio-cultural phenomena.

>>>>

<!-- two types of network analysis -->
Network analysis has been increasingly applied by archaeologists in recent years to deal with past interactions and explore the underlying mechanisms. There are two common approaches to characterize network properties at two distinct scales: node/edge level and graph level [@Peeples2019]. Node level focuses on the role of node in a network, such as centrality, representing the individual influence or social prominence in a group, while graph level assesses the whole network attributes, such as density, clustering in a network, to generalize relationship patterns [@Mills2017; @Peeples2019]. By quantifying those network properties, archaeologists can answer a wide range of research questions. Examples includes exploring the political centralization in the Kofun period in Japan through the hierarchical communication network constructed by prestige goods [@Mizoguchi2013], the investigation of long term inter-site relationships from the Epipalaeolithic to the early Neolithic in the Near East according to trade items [@Coward2013]. Regarding burials, @Sosna2013 examined spatial pattern of burials from the Early Bronze Age in Rebesovice with two hypothesized networks constructed according to cultural and chronological similarity between burials. Recently, complex network modeling evaluates networks at both node and graph level through simulations of particular processes and statistically testing the formation of network properties [@Brughmans2013; @Brughmans2018; @Freeman2004; @Salvini2010]. Such application includes simulations and testing food exchange modes for Ancestral Pueblos on the aggregation of households in the American Southwest simulation [@Crabtree2015], or exploring the diffusion of fired bricks across Hellenistic Europe by comparing similarity networks of sites with random networks. Other examples show the assessment of hunter-gatherer exchange networks structure across the Kuril Islands using bootstrap simulation based on ceramic composition [@Erik2015]. This paper will use a novel Bayesian approach on exponential random graph model to investigate the formation of relationship between burials in northeastern Taiwan to explore the indirect impact of foreign contacts on social organization.

@@

<!-- archaeological specific case study -->
In this study, we explore the changes in social inequality in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan, which covers the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century [@Chen2007]. We assume that the social changes at Kiwulan would be supported by chronological differences in the structure of burial networks. We examine burial networks before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) does the observed burial data after the start of European presence resemble the network of our hypothesized model? (2) did European colonial activities in 17th century Taiwan result in an increased social inequality in an Indigenous society in ways that can be detected by burials networks? (3) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of statistical modeling for burials in understanding the indirect effects of a colonial presence on Indigenous groups. This is important to understand the reactions of Kiwulan residents in the context of indirect impacts of European colonization.

@@

<!-- current approaches to network analysis: important to show how what we are doing is amazing and new and not done before by other archaeologists. We need a sentence or two to say what do archaeologists usually do for network statistics... and make it clear that we are doing something new -->

Bayesian approach to Exponential Random Graph Models (ERGMs) is an effective way to examine relationships by incorporating prior information into networks to better understand dependencies of variables in complex networks and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. Current network analysis used by archaeologist is mostly restricted to a single rational structure without consideration of interaction between network variables. Our use of Bayesian inference on ERGMs is the first application to archaeological data that can bring new insights to understand past social structures by characterizing network properties as a whole. ERGMs are an important family of statistical models for networks that allows direct modeling for the formation of edges, or ties, between nodes [@Robins2007]. The assumption is that possible ties in a network are random variables and dependent on actor variables or the presence or absence of other ties [@Robins2007]. In other words, networks in ERGMs are viewed as dependent variables, where network dependencies and the attributes of nodes/edges can influence the formation of a tie [@Snijders2011]. For example, nodes with similar attributes are more likely to form a relationship, such as friendship between people with the same hobby. Ties form a small structure in a network called a graph configuration, that describes the form of dependence, such as reciprocity (relationship between two actors), transitivity or clustering (relationship between two actors through a shared third actor), homophily (relationship between actors with a similar attribute), and popularity (actors have many relationships with others) [@Robins2007; @Snijders2006; @Morris2008]. Those configurations represent the structure or the property of a network and can be expressed by network statistics. By modeling those network statistics as direct functions of ties by specifying the forms of configurations, we can generate a distribution of random networks that represent our hypothesis-based model [@Morris2008; @Robins2007]. Such distribution consists of a large number of possible networks that enables statistical inference and comparison with an observed network [@Robins2007]. ERGMs helps us understand whether an observed network shows significantly more or less of a property of interest than the random networks generated from our model assumptions.

<!-- ERGM in archaeology, and limitations --> 
To date, ERGM has only been used once in archaeology, by @Amati2019 who reconstructed three different networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore their interaction mechanisms. These network models evaluated different hypotheses about the importance of three variables, proximity, inter-cultural items, and pottery types, on the presence of hub sites they observed. They found that the presence of hubs can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively. However, they also point out some limitations of ERGMs, such as static outcomes and sensitivity to missing data. In addition, it is difficult in ERGMs to estimate model parameters and interpret the result due to intractable likelihood normalizing constants and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a function of the model parameter for making probability distributions integrate to one, which becomes harder to compute with larger set of networks [@Caimo2017bayesian]. This is also termed "doubly intractable" since both the likelihood normalizing constant and the marginal likelihood (the evidence of the posterior) are hard to derive [@Caimo2013; @Lyne2015]. Model degeneracy is another issue that probability models tend to overestimate a small number of extreme graphs by assigning too much weight, such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. One solution to these limitations is available by implementing ERGMs in a Bayesian framework.

@@

<!-- comparison between two networks using bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To understand the robustness of comparison between two networks, we used vertex bootstrap to cross-validate the results based on Bayesian network analysis. Vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) of a network to quantify standard errors in the network statistics of interest [@Chen2019; @Snijders1999]. This enables the evaluation of uncertainty for networks and tests the difference between multiple networks by examining their confidence intervals for the network population. Networks with unoverlapped intervals means there is a significant difference. We set the resampling size to 1,000 networks to compute endogenous network statistics, including density, transitivity, and popularity for our two networks. The result shows a significant difference in network popularity between the two networks, which is consistent with our finding of negative popularity in the pre-European period and positive popularity in the post-European period using the Bayesian approach. For density and transitivity, the 95% confidence intervals from two networks overlapped with each other that indicates no significant difference. This is also consistent with our results of Bayesian modeling, where both networks present similar positive or negative effects some degree of difference.

Using the same vertex bootstrap procedure, we can also test the difference of network populations with a t-test on density, transitivity, and popularity. The results show that there are significant differences in all network statistics. Moreover, we explored the sample size effect by removing nodes at certain percents (5-40%) for both networks and compared their confidence intervals. The results demonstrate a consistent difference between the two networks that means they are robust to the different sizes. In general, those statistical methods support the presence of anthropologically meaningful differences in the pre-and post-European networks. The procedure of cross-validation ensures our results can provide robust comparison between two networks without sample size effects.

@@

<!-- summary of the research by answering questions raised in Intro, and anthropological contribution -->
In this paper, we presented a naval approach for studying burial relations using ERGM network analysis within a Bayesian framework to rigorously examined social changes in a European percicolonial context in northeastern Taiwan. We tested the hypotheses of chronological changes in interburial networks at Kiwulan with the evaluation of both endogenous and exogenous network effects. The results support our hypothesize model that the relationship between burials changed after European colonization period in the 17th century. Before the arrival of the Europeans, the network properties has a tendency of more closed burial relationships with ritual elements as the key formation mechanism. After the European arrival, the network has a tendency of centralization relative to the rarity of goods. The changes in formation mechanisms for networks from ritual-oriented practices to wealth-oriented possession suggest the behavior of wealth accumulation and an increased differentiation stimulated by the European presence and associated long-distance trade network. This aliens with changes in social complexity from a more corporate to a more network mode society that stress different strategies for achieving power and different degrees of social inequality, indicative of an increased social inequality. Using burial data with historical documents, we are able to detect the processes of changes in Indigenous societies in the periphery of European colonial bases to better understand indirect impacts that are underestimated in this region. This study not only discuss pericolonal impacts that are not well understood yet, but also points out the agency of local people to manipulate colonial power instead of being passive recipients of trade goods.

<!-- technical and methodological implication-->
This case study highlights the methodological improvement that Bayesian inference on ERGM can inform and enhance studies in rational data in archaeology. A Bayesian framework can reduce the effects of small sample size or missing data commonly present in archaeological data by incorporating prior information and MCMC estimations. Moreover, probabilistic inference provides robust results of interrelations that are not statistically tested by traditional network approaches. Bayesian network modeling can be applied to a wide range of archaeological data to examine the formation of relationships. This allows in-depth discussion of dynamic processes of relationship formation and their underlying factors that are informative of historical trajectories or socio-cultural phenomena.
